---
title: "1 Discriminant analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
draft: true
execute: 
  eval: false
---

# What is Linear Discriminant Analysis (LDA)?

A definition from [IBM](https://www.ibm.com/docs/en/spss-statistics/beta?topic=features-discriminant-analysis):

Discriminant analysis builds a predictive model for group membership. The model is composed of a discriminant function (or, for more than two groups, a set of discriminant functions) based on linear combinations of the predictor variables that provide the best discrimination between the groups. The functions are generated from a sample of cases for which group membership is known; the functions can then be applied to new cases that have measurements for the predictor variables but have unknown group membership.

# Script `Script_niche1.R`

```{r setup, include=FALSE}
# Install packages
if (!require("pacman")) install.packages("pacman")
if (!require("MASS")) install.packages("MASS")
if (!require("tidyverse")) install.packages("tidyverse")
pacman::p_load(c("MASS", "tidyverse"), character.only = TRUE)
```

Import Data:

```{r}
#this is a way to open a datasheet using a txt file (which can be saved from excel)
lizards<-(read.table("/Users/hu_zhehao/Library/Mobile Documents/com~apple~CloudDocs/Files/Obsidian/Cafe AgX HQ 2.0/004 Notes/M.Sc. Biology/Community Ecology/Practice/transect11_day.txt",header=TRUE))
#to see the data, click on it at the upper pannel on the right or use the command: 
# list(lizards) #all the data
head(lizards) #just the first lines
```

```{r}
summary(lizards)
```

To test if the datas are with equal variances, the test tests of the null that the variances in each of the groups are the same.

```{r}
bartlett.test(lizards$distance + lizards$height + lizards$time ~ lizards$Species, data=lizards)
```

P is way bigger than significance, not rejecting null, they have all equal variance.

Now we do the LDA:

```{r}
fit <- lda(Species ~ distance + height + time, data=lizards, na.action="na.omit", CV=TRUE)
#na.action specifies what to do with variables with missing values. na.omit excludes samples with missing values and proceeds with the analyses.
#if such samples exist but na.action is not specified, the analysis won't run (default option)
# If CV = TRUE the return value is a list with components class, the MAP classification (a factor), and posterior, posterior probabilities for the classes.
fit # show results 
```

- Posterior distribution is the mean value of the simulation.

```{r}
ct <- table(lizards$Species, fit$class)
diag(prop.table(ct, 1))
sum(diag(prop.table(ct)))
```

`sum` will give us the probability of getting a random lizard and assigning it to the correct species only based on its habitat use data #markTodo How?
In another woord, this shows if the species have difference in using habitat, in our case, not. #markTodo but how is this value calculated

Now we do the LDA again, but with `CV=FALSE`

```{r}
model <- lda(Species ~ distance + height + time, data=lizards)
model
plot(model, dimen = 2)
```

Use ggplot for visualization.
