---
title: 3 vegan - diversity, rarefaction, and species richness
date: "`r format(Sys.time(), '%B %d, %Y')`"
draft: true
execute: 
  echo: true
  eval: false
---

```{r setup}
library(tidyverse)
library(dplyr)
library(vegan)
library(readxl)
library(rmarkdown)

setwd("/Users/hu_zhehao/Library/Mobile Documents/com~apple~CloudDocs/Files/Obsidian/Cafe AgX HQ 2.0/004 Notes/M.Sc. Biology/Community Ecology")
community<- read_excel("Practice/Diversity_test_data.xlsx", sheet = "Test")
```

# Raw

`?vegdist`: **Gower**, **Bray--Curtis**, indices are good in detecting underlying ecological gradients (Faith et al. 1987). Why using dissimilarity instead similarity?

our samples: rarefaction set to minimal sample size of 5, a descent analysis will take this out and analyse is again.

---

## Data

We use the data from `Diversity_test_data.xlsx`:

```{r echo=FALSE}
str(community)
paged_table(head(community))
```

## Diversity
### Introduction of diversity index

A diversity index is a mathematical measure of speceis diversity in a given community, it is based on the **species richness** (the number of species present) and the **species abundance** (the number of individuals per species). 

There are two types of diversity indices:

- dominance indices
- information statistic indices

### Shannon Index (H)

$$
H = - \sum^{s}_{i=1}p_i \ln p_i
$$

The **Shannon index** is an information statistic index, which means it assumes all species are represented in a sample and that they are randomly sampled, where $p$ is the proportion ($\frac{n}{N}$) of individuals of one particular species found ($n$) divided by the total number of individuals found ($N$), and $s$ is the number of species.

### Simpson index (D)

$$
D = \frac{1}{ \sum\limits^{s}_{i=1} p_i^2}
$$
The **Simpson index** is a dominance index, meaning it gives more weight to common or dominant species. 


`vegan::diversity()` calculates the ecological diversity indices, with Shannon index being default, Simpson or Invisimpson index can also be choosed.

```{r}
simpson <- diversity(community[,3:11], "simpson") 
simpson 

shannon <- diversity(community[,3:11])
shannon #Typically ranges from 1.5 - 3.4, higher = more diverse 
```

```{r echo=FALSE}
par(mfrow = c(1, 2))  # use par to generate panels with 1 row of 2 graphs
hist(simpson)
hist(shannon)
```

Simpson index shows the probability of randomly drawing two observations given the observations are from different species.

Shannon index gives more importance to rare species, therefore shows higher frequency of the intermediate value.

## Dissimilarity

```{r echo=FALSE}
par(mfrow = c(1, 2))
bray = vegdist(community[,3:11], "bray") #Bray-Curtis does not include joint absences 
gower = vegdist(community[,3:11], "gower") #Gower includes joint absences
hist(bray, xlim = range(0.0,1.0))
hist(gower, xlim = range(0.0,1.0))
```

## Rarefaction, Species richness

```{r}
spAbund <- rowSums(community[,3:11])
spAbund
raremin <- min(spAbund)
raremin
```

```{r}
sRare <- rarefy(community[1,3:11], sample = raremin)
sRare
```



```{r}
require(reshape2)
community$transect <- rownames(community[,3:11]) 

str(melt(community))
#tidyrare <- rarecurve(community[,3:11], tidy = TRUE)
#paged_table(tidyrare)
```


